Nonparametric Clustering Based on Energy Statistics

### 2017-05-18 qcqp_clustering/draft/energy_clustering.tex
figures draft

\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}

@@

\graphicspath{{./figs}}

@@

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}\label{eq:energy}
\Energy(P, Q) = 2 \E \| X - Y\| - \| X - X' \| - \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}\label{eq:energy2}
\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \| X - X' \|^{\alpha} - 
\| Y - Y' \|^{\alpha},
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
equality to zero only requires
equality of the means, and thus it does not imply equality of distributions.

@@

We can formulate an iterative algorithm to find
an approximate solution to \eqref{eq:qcqp2} on the same lines
as kernel $k$-means.
Let $t$ be the iteration time. First precompute the kernel
matrix $G$, fix the number of clusters $k$, then
perform the following steps:
\begin{enumerate}
\item \label{initialize} 
Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
determines the label matrix $\Zt^{(0)}$.
\item \label{assignment} For each datapoint $x_i$ compute
its cluster assignment through
\begin{equation}
\label{eq:algo}
\Zt^{(t+1)}_{ij} = 
\begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
\item If converged return $\Zt^{(t+1)}$, otherwise
set $t = t+1$ and repeat step~\ref{assignment}.
\end{enumerate}
If data is $D$-dimensional, computing $G$ has complexity
$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
iterations,
the total complexity of the algorithm is $O(n^2(D + T))$.
We can initialize the algorithm in step~\ref{initialize} with any method
we want, a good alternative is the initialization from $k$-means++.

@@

Here we consider an alternative approach in the simplest case possible, which
is one-dimensional data and a two-class problem.
If data is one-dimensional and we choose
$\rho(x,y) = |x - y|$, we can actually compute 
\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
a direct solution to \eqref{eq:minimize}. 
This is done by noticing that
\begin{equation}
\begin{aligned}
|x - y|  &= \Ind{x \ge y} (x - y) -
\Ind{x < y} (x - y) \\
&= 
\left( \Ind{x \ge y} - \Ind{x < y} \right) x + 
\left( \Ind{y > x} - \Ind{y \le x} \right) y
\end{aligned}
\end{equation}
where $\Ind{}$ is the indicator function.
Let $\C$ be a partition with
$n$ elements. Using the above distance in \eqref{eq:g_def} we have
\begin{equation}
\label{eq:g_ind}
g(\C,\C) = \dfrac{1}{n^2} \sum_{x \in \C} 
\sum_{y \in \C} 
\left(
\Ind{x \ge y} + \Ind{y > x} - 
\Ind{x \ge y}-\Ind{x < y} \right) x.
\end{equation}
The sum over $y$ can be eliminated since the term in
parenthesis is simply counting the number of elements in $\C$ that satisfy
the conditions of the indicator functions. Therefore, assume
that we order the data as 
$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$. Thus, we
can write \eqref{eq:g_ind} as
\begin{equation}
\label{eq:g1d}
g(\C, \C) = \dfrac{2}{n^2} \sum_{j=1}^n (2j - 1 - n) x_j .
\end{equation}
Note that the cost of computing this is $O(n)$, once the data
is sorted which costs at the most $O(n\log n)$.
Now suppose we order each partition, obtaining $\mathbb{X} = \cup_{j=1}^k
\bar{\C}_j$. Then \eqref{eq:within} is given by
\begin{equation}
\label{eq:w1d}
W(\{ \bar{\C}_i \}) = 2
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{\ell=1}^{n_j} (2\ell - 1 - n_j) x_\ell.
\end{equation}
For a two-class problem, we can cluster the data by computing \eqref{eq:w1d}
for each split and pick the one with minimum value. This is done
as follows:
\begin{enumerate}
\item Sort the pooled data, obtaining the sorted set $\bar{\mathbb{X}}$ with
$n$ elements.
\item For each $t=1,2,\dotsc,n$, make two partitions 
$\bar{\C}_1^{(t)} = [x_i]_{i=1}^t$ and 
$\bar{\C}_2^{(t)} = [x_i]_{i=t+1}^n$. 
\item Compute $W^{(t)} = W(\{ \bar{\C}_1^{(t)},\bar{\C}_2^{(t)}  \})$ 
from \eqref{eq:w1d}.
\item The best split is $t^\star = \argmin_t W^{(t)}$, which
determines the final clusters.
\end{enumerate}
Notice that this procedure does not require any initialization, however,
it only works for one-dimensional data and for Euclidean distance.

@@

>>>>

\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[inline]{enumitem}

@@

\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}

@@

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}\label{eq:energy}
\Energy(P, Q) = 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}\label{eq:energy2}
\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha},
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
equality to zero only requires
equality of the means, and thus it does not imply equality of distributions.

@@

We can formulate an iterative algorithm to find
an approximate solution to \eqref{eq:qcqp2} on the same lines
as kernel $k$-means.
Let $t$ be the iteration time. First precompute the kernel
matrix $G$, fix the number of clusters $k$, then
perform the following steps:
\begin{enumerate}
\item \label{initialize} 
Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
determines the label matrix $\Zt^{(0)}$.
\item \label{assignment} For each datapoint $x_i$ compute
its cluster assignment through
\begin{equation}
\label{eq:algo}
\Zt^{(t+1)}_{ij} = 
\begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
0 & \mbox{otherwise.}
\end{cases}
\nonumber
\end{equation}
\item If converged return $\Zt^{(t+1)}$, otherwise
set $t = t+1$ and repeat step~\ref{assignment}.
\end{enumerate}
If data is $D$-dimensional, computing $G$ has complexity
$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
iterations,
the total complexity of the algorithm is $O(n^2(D + T))$.
We can initialize the algorithm in step~\ref{initialize} with any method
we want, a good alternative is the initialization from $k$-means++.

@@

Here we consider an alternative approach in the simplest case possible, which
is one-dimensional data and a two-class problem.
If we choose
$\rho(x,y) = |x - y|$, we can actually compute 
\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
a direct solution to \eqref{eq:minimize}. 
This is done by noticing that
\begin{equation}
\begin{aligned}
|x - y|  &= \Ind{x \ge y} (x - y) -
\Ind{x < y} (x - y) \\
&= 
\left( \Ind{x \ge y} - \Ind{x < y} \right) x + 
\left( \Ind{y > x} - \Ind{y \le x} \right) y
\end{aligned}
\end{equation}
where $\Ind{}$ is the indicator function.
Let $\C$ be a partition with
$n$ elements. Using the above distance in \eqref{eq:g_def} we have
\begin{equation}
\label{eq:g_ind}
g(\C,\C) = \dfrac{1}{n^2} \sum_{x \in \C} 
\sum_{y \in \C} 
\left(
\Ind{x \ge y} + \Ind{y > x} - 
\Ind{x \ge y}-\Ind{x < y} \right) x.
\end{equation}
The sum over $y$ can be eliminated since the term in
parenthesis is simply counting the number of elements in $\C$ that satisfy
the conditions of the indicator functions. Therefore, assume
that we order the data as 
$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$. Thus, we
can write \eqref{eq:g_ind} in the following very simple form:
\begin{equation}
\label{eq:g1d}
g(\C, \C) = \dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
\end{equation}
Note that the cost of computing this is $O(n)$, once the data
is sorted which costs at the most $O(n\log n)$.
Now suppose we order each partition, obtaining $\mathbb{X} = \cup_{j=1}^k
\bar{\C}_j$. Then \eqref{eq:within} is given by
\begin{equation}
\label{eq:w1d}
W(\{ \bar{\C}_i \}) = 
\sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
\end{equation}
For a two-class problem, we can cluster the data by computing \eqref{eq:w1d}
for each possible split, and pick the one with minimum value. This is done
as follows:
\begin{enumerate}
\item Sort the pooled data, obtaining the sorted set $\bar{\mathbb{X}}$ with
$n$ elements.
\item For each $t=1,2,\dotsc,n$, make two partitions 
$\bar{\C}_1^{(t)} = [x_i]_{i=1}^t$ and 
$\bar{\C}_2^{(t)} = [x_i]_{i=t+1}^n$. 
\item Compute $W^{(t)} = W(\{ \bar{\C}_1^{(t)},\bar{\C}_2^{(t)}  \})$ 
from \eqref{eq:w1d}.
\item The best split is $t^\star = \argmin_t W^{(t)}$, which
determines the final clusters.
\end{enumerate}
Notice that this procedure does not require any initialization, however,
it only works for one-dimensional data and for Euclidean distance.

@@

\begin{figure}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{gauss_means.pdf}\\[-.8em]
(a)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{gauss_covs.pdf}\\[-.8em]
(b)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{gauss_pis.pdf}\\[-.8em]
(c)
\end{minipage}
\caption{
\label{fig:dimension}
gadsga
}
\end{figure}

\begin{figure}
\centering
\caption{
\label{fig:unbalanced}
gadsga
}
\end{figure}

\begin{figure}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{cigar_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{circles_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{spirals_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{cigars.pdf}\\[-.8em]
(a)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{circles.pdf}\\[-.8em]
(b)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{spirals.pdf}\\[-.8em]
(c)
\end{minipage}
\caption{
\label{fig:dimension}
gadsga
}
\end{figure}
